Lattices of acyclic pipe dreams

### 2023-04-20 acyclicFacets/draft/latticePipeDreams.tex
acknoledgments
Added Section 4.5 about the acyclic property for v-Tamari

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}
\newtheorem{notation}[theorem]{Notation}

@@

@@

@@

>>>>

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{openproblem}[theorem]{Open problem}

@@

\vspace*{.5cm}

@@

\subsection{The acyclic property for $\nu$-Tamari lattices}
We would like to close this section stating an open problem that relates our results to the $\nu$-Tamari lattices introduced by Préville-Ratelle and Viennot in~\cite{preville_nu_tamari_2017}. These posets are indexed by a lattice path $\nu$ consisting of a finite number of north (N) and east (E) steps, and coincide with the classical Tamari lattices when $\nu=(NE)^n$.

In~\cite{ceballos_nu_trees_2020}, it was shown that the $\nu$-Tamari lattice can be obtained as the increasing flip poset on pipe dreams~$\pipeDreams(\omega_\nu)$ for an explicit permutation $\omega_\nu$ associated to $\nu$. The permutations of the form~$\omega_\nu$ can be easily characterized as follows. We refer to~\cite{ceballos_nu_trees_2020} for details.

A permutation $\omega$ is called \defn{dominant} if its Rothe diagram, \ie the set $$\set{(\omega(j),i)}{i<j,\ \omega(i)>\omega(j)}$$ in matrix notation, is the Ferrer's diagram of a partition located at the top left corner $(1,1)$. Equivalently, a permutation~$\omega$ is dominant if and only if it is 132-avoiding.

Given a permutation $u$ of $[m]$ and a permutation $v$ of $[n]$, we denote by $u\oplus v$ the permutation of $[m+n]$ defined by 
\[
(u\oplus v)(i)=
\begin{cases}
u(i), & \text{for } 1\leq i\leq m \\
v(i)+m, & \text{for } m<i\leq m+n.
\end{cases}
\] 
The permutations $\omega_\nu$ associated to lattice paths $\nu$ are those of the form~$1\oplus w$ for some dominant permutation $w$, see~\cite{ceballos_nu_trees_2020}.

\begin{openproblem}[The acyclic property for $\nu$-Tamari lattices]
\label{prob_nu_acyclicproperty}
Show that if $\omega$ is a dominant permutation then all pipe dreams with permutation $1\oplus \omega$ are acyclic. %In particular, the corresponding $\nu$-Tamari lattice is the increasing flip poset on acyclic pipe dreams~$\acyclicPipeDreams(1\oplus \omega)$.
\end{openproblem}

This property is known to be true when $\omega=n\dots21$ is the reverse permutation. In this case, the increasing flip poset is the classical Tamari lattice. Removing the vertex 1 from the contact graph of a pipe dream~$P\in \pipeDreams(1\oplus \omega)$ returns the binary tree corresponding to $P$ (with the leaves removed and the internal nodes labeled appropriately with the numbers from 2 to $n+1$). The edges of the binary tree are oriented going away from the root of the tree.  
We have also checked this property computationally for all dominant permutations up to $n=8$.

Applying our Theorem~\ref{thm:pipeDreamQuotient}, we get the following consequence.

\begin{proposition}
If the statement of Open Problem~\ref{prob_nu_acyclicproperty} holds, then the $\nu$-Tamari lattice is a lattice quotient of the interval~$[e,\omega_\nu]$.
\end{proposition}

%\label{thm:pipeDreamQuotient}
%The Hasse diagram of the lattice quotient~$[e,\omega]/{\equiv_\omega}$ is isomorphic to the increasing flip graph on~$\acyclicPipeDreams(\omega)$.
%Hence, the transitive closure of the increasing flip graph on~$\acyclicPipeDreams(\omega)$ is a lattice.

@@

\section*{Acknowledgments}

We thank Florent Hivert for various discussions.
